A comprehensive Ancestral Recombination Graph for SARS-CoV-2 inferred in real time

### 2025-09-25 latex/main.tex
Updates from Overleaf

[FIXME - discuss BA.5 and the BA.2 event instead.]
Our results suggest that Delta and Omicron BA.2 might have recombinant origins,
but we interpret these results with caution,
as these VOCs are undersampled around the time of their emergence (REF).
Consistent with this, the presence of mutations and conflicting loci matching non-focal parents
observed in the copying patterns of these VOCs (Figure 3D) suggests the existence of
many genetically distinct, unsampled ancestors.
Future work is needed to test whether these VOCs descended from recombination events,
with improved sampling, if possible.

@@

We hypothesise that there are a set of samples
from a number of Pango lineages (including B.1.627, B.1.631, B.1.634)
which, should they be added to the ARG, would break up the long branch,
with XB being a recombination between two of them.
In the absence of these, sc2ts cannot detect XB as a recombinant.
Also, note that the XB consensus mutations all occur on or above the long branch.
We suspect that the child node of the long branch
represents the (potentially recombinant) origin of XB, and
the B.1 samples at the bottom left of the subgraph
may be misclassified by pangolin.

@@

\subsubsection{XW}
All the XW samples descend from a well supported recombination node.
Its left parent is a BA.1.1.15-labelled sample (ERR8069658), and
its right parent is a BA.2-labelled sample (ERR9086669).

@@

>>>>

[FIXME - discuss BA.5 (kept as a recombination event but only because we can't 
rewire it post-hoc, owning to subsequent tree building) and the BA.2 event (doesn't
pass QC.]
[Our results suggest that Delta and Omicron BA.2 might have recombinant origins [no longer true],
but we interpret these results with caution,
as these VOCs are undersampled around the time of their emergence (REF).]
Note that the presence of many de-novo mutations
observed in the copying patterns of XBB (Figure 3D) suggests the existence of
many genetically distinct, unsampled ancestors [FIXME: in this case, probably BJ.1, see
https://github.com/jeromekelleher/sc2ts-paper/issues/336].
Future work is needed to test whether these VOCs descended from recombination events,
with improved sampling, if possible.

@@

We hypothesize that there are a set of samples
from a number of Pango lineages (including B.1.627, B.1.631, B.1.634)
which, should they be added to the ARG, would break up the long branch,
with XB being a recombination between two of them.
In the absence of these, sc2ts cannot detect XB as a recombinant.
Also, note that the XB consensus mutations all occur on or above the long branch.
We suspect that the child node of the long branch
represents the (potentially recombinant) origin of XB, and
the B.1 samples at the bottom left of the subgraph
may be misclassified by pangolin.

@@

\subsubsection{XW}
All the XW samples descend from a well supported recombination node.
Its left parent is a BA.1.1.15-labeled sample (ERR8069658), and
its right parent is a BA.2-labeled sample (ERR9086669).

@@

\subsubsection{Xx: XZ/XAC/XAD/XAE/XAP}
WIP

\subsubsection{XAF}
https://github.com/jeromekelleher/sc2ts-paper/discussions/638

Of the 21 XAF samples in the Viridian v04 dataset,
only XAF sample passed our QC filters and was added to the ARG:
many have missing Ns in their consensus sequences (> 500) or
require a large number of mutations to incorporate into the ARG
(i.e. have a high HMM cost).
However, we find a small recombinant BA1.1/BA.2 clade of 37 samples,
associated with the BA.2-labeled sample ERR9755541,
within which the single XAF sample is nested.
The inferred parents and breakpoint at 10,447
aligns with that proposed for XAF
(https://github.com/cov-lineages/pango-designation/issues/676).
Inspection of the omitted XAF samples with high HMM cost shows that
they would attach as close sisters to the included XAF sample,
albeit on branches with many new mutations.
Hence we believe the current XAF topology in the ARG is reasonable,
although somewhat tentative, given the number of included samples.

The mutation-rich branch in the subgraph with
a 5-base deletion (11283-11287) and insertion (11292-11296)
is likely to be a simple deletion alignment error.

\subsubsection{XAJ}
https://github.com/jeromekelleher/sc2ts-paper/discussions/634

In the Pango designation issue
(https://github.com/cov-lineages/pango-designation/issues/826),
XAJ is suggested to be a recombinant between
BA.2.12.1 (left parent) and BA.4 (right parent).
XAJ was detected as a non-recombinant by sc2ts.
There are only three loci supporting
the right BA.4 parent in the ARG (T22917G, T23018G, and G23040A),
which means that XAJ fell below our threshold of detection.
There is also a deletion (H69 in Spike) supporting
the BA.4 parent (see the Pango designation issue),
but it was treated as missing data during ARG inference.
If this deletion were accounted for,
XAJ should have been detected as a recombinant.

\subsubsection{XAN/XAV}
https://github.com/jeromekelleher/sc2ts-paper/discussions/783
https://github.com/jeromekelleher/sc2ts-paper/discussions/785

XAN and XAV are shown in this subgraph, as they share some mutations.
In the Pango designation issues
(https://github.com/cov-lineages/pango-designation/issues/771;
https://github.com/cov-lineages/pango-designation/issues/911),
XAN and XAV are suggested to be a recombinant between
BA.2 (left parent) and BA.5 (right parent).
Both these Pango lineages were detected as non-recombinants by sc2ts,
because the number of supporting loci fell below our threshold of detection.
In the ARG, we found only one mutation (C9866T)
supporting the suggested BA.2 parent for both XAN and XAV.

\subsubsection{XAS}
https://github.com/jeromekelleher/sc2ts-paper/discussions/784

XAS was proposed as a recombination between
BA.5 (left parent) and BA.2 (right parent).
However, we find that this is only two mutations
different from e.g. ERR9929846, which is a BA.4 sample.
In fact, XAS only differs from the BA.4 root by a single mutation,
C27945T, which was identified in the original Pango designation issue
(https://github.com/cov-lineages/pango-designation/issues/882)
as an important ORF truncation mutation
which is also present in a few BA.2.65 samples.
It is possible that C27945T was acquired by recombination,
but this is clearly below our threshold of detection.

\subsubsection{XAZ}
https://github.com/jeromekelleher/sc2ts-paper/discussions/786

In the Pango designation issue
(https://github.com/cov-lineages/pango-designation/issues/797),
XAZ is suggested as a two-breakpoint recombinant between
BA.2.5 and BA.5 (the middle parent),
with three sites supporting the BA.2.5 parent
on the left end (C2232T, C3317T, and T3358C).
In the ARG, this is very close to being a recombinant of
BA.2.5 (ERR9615610) on the left side of the breakpoint,
as not only are the three mutations present in this sample,
but there is another site which supports recombination 
C1912T, which is not reported in the Pango designation issue).
XAZ has C44T which is lacking in all BA.2.5 samples,
leaving only three net supporting loci for a putative recombinant.
However, position 44 is often clipped,
so this is not strong evidence against a recombinant, and
therefore XAZ is a marginal case of recombination.
We found no evidence for the third segment
inherited from a BA2.5 parent on the right end.
In the ARG, all the XAZ samples form a clade.
Here we display only the first 20 XAZ samples.

\subsubsection{XBB}
WIP

\subsubsection{XBB.1}
WIP

\subsubsection{XBD}
All the XBD samples descend from a well supported recombination node.
Its left and right parents are reconstructed internal nodes,
both of which are directly supported by samples
(SRR21382561 and ERR9825609, respectively)
that are identical to them.

\subsubsection{XBE}
WIP

\subsubsection{XBF}
All the XBF samples descend from a well supported recombination node,
with three mutations on the left of the breakpoint.
Its left parent is a sample assigned to BA.5.2.1 (SRR20305553).
Its right parent is a reconstructed internal node
that is supported by a CJ.1-labeled sample (SRR21948068),
which differs by three mutations.

\subsubsection{XBG}
All the XBG samples descend from a well supported recombination node,
with two mutations, one on each side of the breakpoint.
Both its left and right parents are samples
(SRR19920084 and ERR9852351, respectively).

\subsubsection{XBH}
All the XBH samples descend from a well supported recombination node,
despite two reversions (T222C and T12789C, respectively),
which occur at excluded sites and
were remapped by parsimony during postprocessing.
If these two sites had been used during ARG inference,
the branch attached to BA.2.1 would have been attached to
the BA.2.5 shown on the top left instead,
leading to a more parsimonious local topology
with two mutations on the left branch.

\subsubsection{XBQ/XBK/XBK.1/CJ.1.3}
WIP

\subsubsection{XBM}
All the XBM samples descend from a well supported recombination node.
Its left parent is a reconstructed internal node
that is directly supported by a BA.2.76-labeled sample identical to it.
Its right parent is a BF.3-labeled sample.

\subsubsection{XBR}
This one XBR sample descends from a well supported recombination node.
Its left parent is a BN.3.1-labeled sample, and
its right parent is a reconstructed internal node
that is directly supported by a BQ.1.25-labeled sample identical to it.
